---
title: "Diachromatic motif analysis"
author: "Jonas Ibn-Salem"
date: "5 März 2019"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Background and aim

Long-range chromatin interactions result in read-pairs mapping in two distinct configurations, simple and twisted. 
Each pair of restriction fragments is classified as indirected or directed depending on similar fraction of simple and twisted read pairs or one outweighing the other.

### Goal

 - [x] Analyze distribution of twisted vs. simple reads to find classification in directed and undirected
 - [x] Analyze CTCF motif distribution 
 - [ ] Analyze all TF motif distributions

### TODO

 - [ ] Analyze other interaction datasets
 - [ ] Analyze other TF motifs
 - [ ] Test if distance is significantly different between interaction classes
 - [ ] Check which score (score_1 or score_2) is the -log_10 p-value of motif hit
 
### Questions to Peter

- Coordinnates of interactions are hg19?
- Coordinates zero-based or one-based?
- Is there a BED file with all restriction fragments?

### Data

We start to analyze the following data sets:

sample              |id          |
--------------------|------------|
Hi-C_untreated_rep1 | SRR5633682 | 
Hi-C_untreated_rep1 | SRR5633683 | 


## Data and parameters

```{r}
require(InteractionSet) # infrastructure for chromatin interactions
require(rtracklayer) # for liftOver of genomic coordinates
require(R.utils)   # to unzip .gz files
require(tidyverse)  # to analyze tabular data
require(fs)         # to work with files and paths 

```

### Set parameters

```{r}
ssr_ids <- c("SRR5633682", "SRR5633683")
# ssr_ids <- c("SRR5633682")

# CTCF motif (JASPAR ID MA0139.1)
CTCF_motif_url_template <- "http://expdata.cmmt.ubc.ca/JASPAR/downloads/UCSC_tracks/2018/mm10/tsv/MA0139.1.chr<N>.tab.gz"
mm10_to_mm9_chain_url <- "http://hgdownload.cse.ucsc.edu/goldenpath/mm10/liftOver/mm10ToMm9.over.chain.gz"

mm10_chroms <- c(as.character(1:19), "M", "X", "Y")

mm10_seq_info <- Seqinfo(genome = "mm10")

RUN_CLEAN <- TRUE

```


```{r}

sample_tab <- tibble(
  id = ssr_ids,
  file = dir_ls(fs::path("data", "Diachromatic"), glob = "*.tsv.gz") %>% 
    str_subset(pattern = ssr_ids)
)


# read interaction files
read_interactions <- function(infile){
  
  df <- read_tsv(infile, col_names = 
                 c("chr_a", "start_a", "end_a", "type_a", "chr_b", "start_b", "end_b", "type_b", "simple_twisted")) %>% 
    separate(simple_twisted, into = c("simple", "twisted"), sep = ":") %>% 
    mutate_at(c("simple", "twisted"), list(parse_integer))
}

```

## Read data

```{r read interactions}

if (RUN_CLEAN) {

  # read each file
  sample_tab <- sample_tab %>% 
    mutate(df = map(file, read_interactions))
  
  write_rds(sample_tab, "sample_tab.rds")

} else {
  sample_tab <- read_rds("sample_tab.rds")  
}

# combine into a single data set
df <- sample_tab %>% 
  unnest(df)

df <- df %>% 
  mutate(
    dist = start_b - end_a,
    simple_twisted_lr = log2( (simple + 1) / (twisted + 1)),
    class = case_when(
      simple_twisted_lr > 1 ~ "simple",
      simple_twisted_lr < -1 ~ "twisted",
      TRUE ~ "undirected"),
    k = simple + twisted,
    k_group = case_when(
      k >= 10 ~ "k >= 10",
      k >= 5 ~ "k >= 5",
      k >= 4 ~ "k >= 4",
      k >= 3 ~ "k >= 3",
      k >= 2 ~ "k >= 2",
    ) %>% 
      as.factor() %>% 
      fct_relevel("k >= 10", after = Inf)
  )

```


### Get CTCF motif location

We download CTCF motif predictions from JASPAR database. 
The CTCF moitf and frequency matrix can be optained from here: http://jaspar.genereg.net/matrix/MA0139.1/.


```{r read motifs}

# 
if (RUN_CLEAN) {
  
  chrom_df <- tibble(
    chr = mm10_chroms,
    motif_url = map_chr(mm10_chroms, ~str_replace(CTCF_motif_url_template, "<N>", .x)),
    df = map(motif_url, read_tsv,
             col_names = c("start", "strand", "score_1", "score_2"),
             col_types = "dcdd")
  )
  
  motif_df <- chrom_df %>% 
    select(-motif_url) %>% 
    unnest(df)

  write_rds(motif_df, "motif_df.rds")

} else {
  motif_df <- read_rds("motif_df.rds")  
}

motif_gr <- GRanges(str_c("chr", motif_df$chr),
                    IRanges(motif_df$start, motif_df$start + 19),
                    strand = motif_df$strand,
                    select(motif_df, score_1, score_2),
                    seqinfo = mm10_seq_info
  )

# get liftover chain

# download and extract chain file
if (RUN_CLEAN) {
  
  if (file_exists("data/mm10ToMm9.over.chain.gz"))
    fs::file_delete("data/mm10ToMm9.over.chain.gz")
  
  if (file_exists("data/mm10ToMm9.over.chain"))
    fs::file_delete("data/mm10ToMm9.over.chain")
  
  download.file(mm10_to_mm9_chain_url, "data/mm10ToMm9.over.chain.gz")
  R.utils::gunzip("data/mm10ToMm9.over.chain.gz")
}
ch <- import.chain("data/mm10ToMm9.over.chain")

# liftover motifs from mm10 to mm9
motif_gr <- rtracklayer::liftOver(motif_gr, ch) %>% 
  unlist()


```

```{r}
ggplot(motif_df, aes(x = score_2)) +
  geom_histogram() + 
  geom_vline(xintercept = 500, color = "red") +
  theme_bw()

```


<!-- We filter CTCF motif locations for motif hits with a significance score corresponding to at least $p = 10^-5$. -->

```{r}
# motif_gr <- motif_gr[motif_gr$score_2 >= 500]
```

This result in `r length(motif_gr)` motifs.

## Analyze interaction types

### Count interactions per file

```{r}
df %>% 
  count(id) %>% 
  ggplot(aes(x = id, y = n, label = n)) + 
    geom_bar(stat = "identity") + 
    geom_text(vjust = "bottom") + 
    # coord_flip() +
    theme_bw() + 
  labs(y = "number of interactions")

```

### Plot simple and twisted counts

```{r}

df %>% 
  sample_n(10^6) %>% 
  ggplot(aes(x = simple, y = twisted)) + 
    geom_bin2d() + 
    facet_grid(. ~ id) +
    scale_fill_distiller(direction = 1) +
    theme_bw() +
    theme(legend.position = "bottom")

df %>% 
  filter(simple <= 20, twisted <= 20) %>% 
  sample_n(10^6) %>% 
  count(id, simple, twisted) %>% 
  ggplot(aes(x = simple, y = twisted, fill = n, label = n)) + 
    geom_bin2d(stat = "identity") + 
    geom_text(size = 3) +
    facet_grid(. ~ id) +
    scale_fill_distiller(direction = 1) +
    theme_bw() +
    theme(legend.position = "bottom")

# df %>% 
#   sample_n(10^6) %>% 
#   ggplot(aes(x = simple, y = twisted)) + 
#     geom_bin2d() + 
#     facet_grid(. ~ id) +
#     scale_x_log10() + 
#     scale_y_log10() +
#     scale_fill_distiller(direction = 1) +
#     theme_bw()

```


### Ratio simple vs. twisted (without zero counts)

```{r}

lfc_df <- df %>% 
  filter(twisted > 0, simple > 0) %>% 
  sample_n(10^6) %>%
  mutate(lfc = log2(simple / twisted))

lfc_df %>%
  ggplot(aes(x = lfc)) +
    geom_histogram(bins = 30) +
    facet_grid(. ~ id) +
    theme_bw() + 
    labs(x = "log2(simple / twisted)")

```

### Ratio simple vs. twisted (with pseudo counts)

```{r}
df %>% 
  sample_n(10^6) %>%
  ggplot(aes(x = simple_twisted_lr)) +
    geom_histogram(bisn = 30) +
    geom_vline(xintercept = c(-1, 1), color = "red") +
    facet_grid(. ~ id) +
    theme_bw() + 
    labs(x = "log2((simple + 1) / (twisted + 1))")

```

### Classification of interactions

We classify interactions based on simple and twisted reads into the folloing calses:

 - *simple*, if log2( (simple + 1) / (twisted + 1)) > 1
 - *twisted*, if log2( (simple + 1) / (twisted + 1)) < -1
 - *undirected*, otherwise

This lets to the follwing number of interactions per class

```{r}
df %>% 
  count(id, class) %>% 
  ggplot(aes(x = class, y = n, label = n, fill = class)) + 
  geom_bar(stat = "identity") +
  geom_text(vjust = "top") +
  facet_grid(. ~ id) +
  theme(legend.position = "none") +
  theme_bw()
```

## Analysze genomic distance

```{r}
df %>% 
  ggplot(aes(x = dist)) +
    geom_histogram(bins = 30) + 
    facet_grid(. ~ id) +
    scale_x_log10() +
    theme_bw() +
    labs(x = "Distance in bp")

df %>% 
  sample_n(1000) %>% 
  ggplot(aes(x = simple_twisted_lr, y = dist)) +
    geom_bin2d() +
    facet_grid(. ~ id) +
    scale_y_log10() +
    scale_fill_distiller(direction = 1) +
    theme_bw()

```

### Distance by class

```{r}
df %>% 
  sample_n(10000) %>%
  ggplot(aes(x = class, y = dist, color = class)) + 
    geom_boxplot() + 
    facet_grid(. ~ id) +
    scale_y_log10() +
    theme_bw()

```


## Analyse motifs

### Convert to InteractonSet object

```{r}
# get regions as GenomicRanges objects
gr_a <- GRanges(df$chr_a, IRanges(df$start_a, df$end_a))
gr_b <- GRanges(df$chr_b, IRanges(df$start_b, df$end_b))

# gi <- GInteractions(gr_a, gr_b, metadata = as.list(select(df, simple, twisted, simple_twisted_lr)))
```


### Motif overlap

We use pre-calculated genomic locations of the CTCF recognition motif from JASPAR database and calculate motif occurrence with restriction fragments. For each interaction, we count the number of motif occurrence in the positive and negative strand for the upstream (a) and downstream (b) anchor fragment.  

```{r}
# split into motifs on positive and negative strand
motifs_pos <- motif_gr[strand(motif_gr) == "+"]
motifs_neg <- motif_gr[strand(motif_gr) == "-"]

# count motifs on A and B anchor of interactions
# and add motif counts to interaction data

df <- df %>% 
  mutate(
    motif_a_pos = countOverlaps(gr_a, motifs_pos),
    motif_a_neg = countOverlaps(gr_a, motifs_neg),
    motif_b_pos = countOverlaps(gr_b, motifs_pos),
    motif_b_neg = countOverlaps(gr_b, motifs_neg)
)

# add convergetn column
df <- df %>% 
  mutate(
    convergent = ifelse(motif_a_pos > 0 & motif_b_neg > 0, "convergent", "not convergent")
  )

```

### Motifs per fragment

```{r}
motif_counts <- df %>% 
  sample_n(10^6) %>% 
  gather("motif_type", "motif_count", motif_a_pos, motif_a_neg, motif_b_pos, motif_b_neg) %>% 
  filter(motif_count <= 10) %>% 
  count(id, motif_type, motif_count)

ggplot(motif_counts, aes(x = motif_count, y = n, fill = motif_type)) + 
  geom_bar(stat = "identity") + 
  # scale_x_log10() + 
  facet_grid(motif_type ~ id) + 
  theme_bw() +
  theme(legend.position = "none")

```


### Analyze motifs by class

```{r}
df %>% 
  group_by(id, class, convergent) %>% 
  summarize(n = n()) %>% 
  mutate(percent_convergent = 100 * n / sum(n)) %>% 
  ggplot(aes(x = class, y = percent_convergent, fill = convergent, label = round(percent_convergent, 2))) +
    geom_bar(stat = "identity") + 
    geom_text(position = position_stack(.5)) +
    facet_grid(. ~ id) +
    coord_flip() + 
    theme_bw() + 
    theme(legend.position = "bottom")

```

We further subdivide interactions according to the total numberof reads k.


```{r}
df %>%
  count(k_group) %>% 
  ggplot(aes(x = k_group, y = n, label = n)) + 
  geom_bar(stat = "identity") + 
  geom_text(vjust = "bottom") +
  theme_bw()

df %>% 
  group_by(id, class, k_group, convergent) %>% 
  summarize(n = n()) %>% 
  mutate(percent_convergent = 100 * n / sum(n)) %>% 
  ggplot(aes(x = class, y = percent_convergent, fill = convergent, label = round(percent_convergent, 2))) +
    geom_bar(stat = "identity") + 
    geom_text(position = position_stack(.5)) +
    facet_grid(k_group ~ id) +
    coord_flip() + 
    theme_bw() + 
    theme(legend.position = "bottom")

```


There is a slight trend towards more convergent oriented CTCF motif pairs for interactions of the class simple, compared to undirected and twisted. This effect is stronger for interactions with higher number of total reads.


```{r}
df %>% 
  group_by(id, class, k_group, convergent) %>% 
  summarize(n = n()) %>% 
  mutate(percent_convergent = 100 * n / sum(n)) %>% 
  filter(convergent == "convergent") %>% 
  ggplot(aes(x = class, y = percent_convergent, fill = class, label = round(percent_convergent, 2))) +
    geom_bar(stat = "identity") + 
    geom_text(vjust = 1.5) +
    facet_grid(k_group ~ id) +
    # coord_flip() + 
    theme_bw() + 
    theme(legend.position = "none") + 
  labs(x = "Interaction class", y = "Percent of interactions with convergent CTCF motifs")

```

